Description of the power system model \emph{medea}

### 2019-09-06 doc/mathematical_description.tex
further consistency improvements in math documentation

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[fleqn]{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\DeclareMathOperator*{\argmax}{arg\,max}

@@

\begin{longtable}{p{3.6cm} c c p{6cm}}
%\caption{Sets} \\ 
\hline
\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{elements} \\
\hline \hline

market zones & $z \in Z$ & \texttt{z} & \texttt{AT, DE} \\ \hline

time periods (hours) & $t \in T$ & \texttt{t} & \texttt{t1, t2, \ldots, t8760} \\ \hline

power generation technologies & $i \in I$ & \texttt{tec} & 
\texttt{nuc, lig\_stm, lig\_stm\_chp, lig\_boa, lig\_boa\_chp,coal\_sub, coal\_sub\_chp, coal\_sc, coal\_sc\_chp, coal\_usc, coal\_usc\_chp, coal\_igcc, ng\_stm, 
ng\_stm\_chp, ng\_ctb\_lo, ng\_ctb\_lo\_chp, ng\_ctb\_hi, ng\_ctb\_hi\_chp, ng\_cc\_lo, ng\_cc\_lo\_chp, ng\_cc\_hi, ng\_cc\_hi\_chp, ng\_mtr, ng\_mtr\_chp, 
ng\_boiler\_chp, oil\_stm, oil\_stm\_chp, oil\_ctb, oil\_ctb\_chp, oil\_cc, oil\_cc\_chp, bio, bio\_chp, heatpump\_pth} \\ \hline

CHP technologies & $j \in J \subset I$ & \texttt{tec\_chp} & \texttt{lig\_stm\_chp, lig\_boa\_chp, coal\_sub\_chp, coal\_sc\_chp, coal\_usc\_chp, ng\_stm\_chp, 
ng\_ctb\_lo\_chp, ng\_ctb\_hi\_chp, ng\_cc\_lo\_chp, ng\_cc\_hi\_chp, ng\_mtr\_chp, ng\_boiler\_chp, oil\_stm\_chp, oil\_ctb\_chp, oil\_cc\_chp, bio\_chp} \\ \hline

power to heat technologies & $h \in H \subset I$ & \texttt{tec\_pth} & \texttt{heatpump\_pth}  \\ \hline

storage technologies & $k \in K$ & \texttt{tec\_strg} & \texttt{res\_day, res\_week, res\_season, psp\_day, psp\_week, psp\_season, battery} \\ \hline

intermittent generators & $n \in N$ & \texttt{tec\_itm} & \texttt{wind\_on, wind\_off, pv, ror}\\ \hline

fuels & $f \in F$ & \texttt{f} & \texttt{nuclear, lignite, coal, gas, oil, biomass, power} \\ \hline

feasible operation region limits & $l \in L$ & \texttt{l} & \texttt{l1, l2, l3, l4} \\ \hline

energy products & $m \in M$ & \texttt{prd} & \texttt{el, ht} \\ \hline \hline
\end{longtable}

@@

\begin{longtable}{p{4.1cm} c l c}
\hline
	\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{unit} \\ \hline \hline
% LOWER-CASE GREEK LETTERS
	distance between countries & $\delta_{z,zz}$ & \texttt{KM(z,zz)} & km \\ \hline
	fuel emission intensity & $\varepsilon_{f}$ & \texttt{EMISSION\_INTENSITY(f)} & $\text{t}_{\ce{CO2}}$ / MWh \\ \hline
	power plant efficiency & $\eta_{i,p,f}$ & \texttt{EFFICIENCY(tec,prd,f)} & MWh / MWh \\ \hline
	efficiency power out & $\eta^{out}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'efficiency\_out')}} &  \\ \hline
	efficiency power in & $\eta^{in}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'efficiency\_in')}} &  \\ \hline	
	value of lost load & $\mu_{z}$ & \texttt{VALUE\_NSE(z)} & \EUR / MWh \\ \hline
	reservoir inflows & $\rho_{z,t,k}$ & \makecell[l]{\texttt{RESERVOIR\_INFLOWS} \\ \texttt{(z,t,tec\_strg)}} & MW \\ \hline	
	minimal conventional generation & $\sigma_{z}$ & \texttt{ANCIL\_SERVICE\_LVL(z)} & GW \\ \hline
	intermittent generation profile & $\phi_{z,t,n}$ & \texttt{GEN\_PROFILE(z,t,tec\_itm)} & $[0,1]$ \\ \hline
	inputs of feasible operating region & $\chi_{i,l,f}$ & \texttt{FEASIBLE\_INPUT(tec,l,f)} & $[0,1]$ \\ \hline
	output tuples of feasible operating region & $\psi_{i,l,p}$ & \texttt{FEASIBLE\_OUTPUT(tec,l,prd)} & $[0,1]$ \\ \hline
% UPPER-CASE LATIN LETTERS	
	capital cost of intermittent generators (specific, annuity) & $C^{r}_{z,n}$ & \texttt{INVESTCOST\_ITM(z,tec\_itm)} & k\EUR / GW \\ \hline
	capital cost of thermal generators (specific, annuity) & $C^{g}_{z,i}$ & \texttt{INVESTCOST\_THERMAL(z,tec)} & k\EUR / GW \\ \hline
	capital cost of storages - power (specific, annuity) & $C^{s}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'cost\_power')}} & k\EUR / GW \\ \hline
	capital cost of storages - energy (specific, annuity) & $C^{v}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'cost\_energy')}} & k\EUR / GW \\ \hline
	capital cost of transmission capacity & $C^{x}_{z}$ & \texttt{INVESTCOST\_ATC} & k\EUR / GW \\ \hline
	energy demand & $D_{z,t,p}$ & \texttt{CONSUMPTION(z,t,prd)} & GW \\ \hline
	initial capacity of thermal generators & $\bar{G}_{z,i}$ & \texttt{INSTALLED\_CAP\_THERM(z,tec)} & GW \\ \hline
	quasi-fixed O\&M cost & $O^{q}_{i}$ & \texttt{OM\_FIXED\_COST(tec)} & k\EUR / GW \\ \hline
	variable O\&M cost & $O^{v}_{i}$ & \texttt{OM\_VAR\_COST(tec)} & \EUR / MWh \\ \hline
	\ce{CO2} price & $P^{e}_{t,z}$ & \texttt{PRICE\_CO2(t,z)} & \EUR / $\text{t}_{\ce{CO2}}$ \\ \hline
	fuel price & $P_{t,z,f}$ & \texttt{PRICE\_FUEL(t,z,f)} & \EUR / MWh \\ \hline
	initial capacity of intermittent generators & $\bar{R}_{z,n}$ & \texttt{INSTALLED\_CAP\_ITM(z,tec\_itm)} & GW \\ \hline
\pagebreak
%\caption[Parameters]{(continued)}\\ 
\hline
	% UPPER-CASE LATIN LETTERS (CONTINUED)
	\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{unit} \\ \hline \hline
	max power out & $\bar{S}^{out}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'power\_out')}} & GW \\ \hline
	max power in & $\bar{S}^{in}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'power\_in')}} & GW \\ \hline
	max energy stored & $\bar{V}_{z,k}$ & \makecell[l]{\texttt{STORAGE\_PROPERTIES} \\ \texttt{(z,tec\_strg,'energy\_max')}} &  \\ \hline
	installed available transmission capacity & $\bar{X}_{z,zz}$ & \texttt{ATC(z,zz)} & GW \\ \hline \hline	
\end{longtable}

@@

\begin{longtable}{p{4.1cm} c l c}
%\caption{Variables}\label{variables} \\ 
\hline
\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{Unit} \\ \hline \hline
fuel burn for energy generation & $b_{z,t,i,f}$ & \texttt{q\_fueluse(z,t,tec,f)} & GW \\ \hline
zonal system cost & $c_{z}$ & \texttt{cost(r)} & k\EUR \\ \hline
fuel cost & $c^{b}_{z,t,i}$ & \texttt{cost\_fuel(z,t,tec)} & k\EUR \\ \hline
emission cost & $c^{e}_{z,t,i}$ & \texttt{cost\_emission(z,t,tec)} & k\EUR \\ \hline
total o\&m cost & $c^{om}_{z,i}$ & \texttt{cost\_om(z,tec)} & k\EUR \\ \hline
capital cost of generators & $c^{g}_{z}$ & \texttt{cost\_invgen(z)} & k\EUR \\ \hline
total cost of non-served load & $c^{q}_{z}$ & \texttt{cost\_nse(z)} & k\EUR \\ \hline
capital cost of intermittent generators & $c^{r}_{z}$ & \texttt{cost\_invitm(z)} & k\EUR \\ \hline
capital cost of storages & $c^{s,v}_{z}$ & \texttt{cost\_invstrg(z)} & k\EUR \\ \hline
capital cost of interconnectors & $c^{x}_{z}$ & \texttt{cost\_gridexpansion(z)} & k\EUR \\ \hline
\ce{CO2} emissions & $e_{z}$ & \texttt{emissions(z)} & t \ce{CO2} \\ \hline
added capacity of conventionals & $\bar{g}^{+}_{z,i}$ & \texttt{invest\_thermal(z,tec)} & GW \\ \hline
decommissioned capacity of conventionals & $\bar{g}^{-}_{z,i}$ & \texttt{decommission(z,tec)} & GW \\ \hline
energy generated by conventionals & $g_{z,t,i,p}$ & \texttt{q\_gen(z,t,tec,prd)} & GW \\ \hline
curtailed energy & $q^{+}_{z,t}$ & \texttt{q\_curtail(z,t)} & GW \\ \hline
non-served energy & $q^{-}_{z,t,p}$ & \texttt{q\_nonserved(z,t,prd)} & GW \\ \hline
added capacity of intermittents & $\bar{r}^{+}_{z,n}$ & \texttt{invest\_res(z,tec\_itm)} & GW \\ \hline
electricity generated by intermittents & $r_{z,t,n}$ & \texttt{q\_itm(z,t,tec\_itm)} & GW \\ \hline
added storage capacity (power) & $\bar{s}^{+}_{z,k}$ & \texttt{invest\_storage\_power(z,tec\_strg)} & GW \\ \hline
energy stored in & $s^{in}_{z,t,k}$ & \texttt{q\_store\_in(z,t,tec\_strg)} & GW \\ \hline
energy stored out & $s^{out}_{z,t,k}$ & \texttt{q\_store\_out(z,t,tec\_strg)} & GW \\ \hline
added storage capacity (energy) & $\bar{v}^{+}_{z,k}$ & \texttt{invest\_storage\_energy(z,tec\_strg)} & GWh \\ \hline
storage energy content & $v_{z,t,k}$ & \texttt{storage\_level(z,t,tec\_strg)} & GWh \\ \hline
operating region weight & $w_{z,t,i,l}$ & \texttt{cc\_weights(z,t,tec,l)} & . \\ \hline
added transmission capacity & $\bar{x}^{+}_{z,zz}$ & \texttt{invest\_atc(z,zz)} & GW \\ \hline
electricity net export & $x_{z,zz,t}$ & \texttt{flow(z,zz,t)} & GW \\ \hline \hline
\end{longtable}

@@

\section{Naming system}
\ctable[
	maxwidth = 160mm,
	pos = h]
	{X p{2.0cm} p{2.0cm} p{2.0cm} p{2.0cm} p{2.0cm}}
	{
	\tnote[$\dagger$]{parameter} 
	\tnote[$\ddagger$]{variable}
	}
	{\FL 
	 &  initial capacity\tmark[$\dagger$] & added capacity\tmark[$\ddagger$] & \makecell[l]{decom-\\missioned \\capacity\tmark[$\ddagger$]} & \makecell[l]{specific \\investment \\cost\tmark[$\dagger$]} & dispatch\tmark[$\ddagger$] \ML
	thermal units & $\bar{G}_{z,i}$ & $\bar{g}^{+}_{z,i}$ & $\bar{g}^{-}_{z,i}$ & $C^{g}_{z,i}$ & $g_{z,t,i,p}$ \ML
	\makecell[l]{intermittent \\units} &  $\bar{R}_{z,n}$ & $\bar{r}^{+}_{z,n}$ & $\bar{r}^{-}_{z,n}$ & $C^{r}_{z,n}$ & $r_{z,t,n}$ \ML
	\makecell[l]{storages \\(power)} & $\bar{S}_{z,k}$ & $\bar{s}^{+}_{z,k}$ & $\bar{s}^{-}_{z,k}$ & $C^{s}_{z,k}$ & $s_{z,t,k}$ \ML
	\makecell[l]{storages \\(energy)} & $\bar{V}_{z,k}$ & $\bar{v}^{+}_{z,k}$ & $\bar{v}^{-}_{z,t,k}$ & $C^{v}_{z,k}$ & na \ML
	transmission &  $\bar{X}_{z,zz}$ & $\bar{x}^{+}_{z,zz}$ & $\bar{x}^{-}_{z,zz}$ & $C^{x}_{z,zz}$ & $x_{z,zz,t}$ \LL
}

@@

\paragraph{Model objective}
\emph{medea} minimizes total system cost $c$, i.e. the total cost of generating electricity and heat from technologies and capacities adequate to meet demand, over a large number  of decision variables, essentially representing investment and dispatch decisions in each market zone $z$ of the modelled energy systems.
\begin{align}
\min c = \sum_{z} (c_{z})
\end{align}
Zonal system costs $c_{z}$ are the sum of fuel cost $c^{b}_{z,t,i}$, emission cost $c^{e}_{z,t,i}$, operation and maintenance cost $c^{om}_{z,i}$, capital costs of investment in conventional and intermittent generation ($c^{g}_{z}$, $c^{r}_{z}$), storage ($c^{s,v}_{z}$) and transmission ($c^{x}_{z}$) equipment, and the cost of non-served load ($c^{q}_{z}$) that accrues when demand is not met, e.g. when there is a power outage. 
\begin{align}
%\begin{split}
c_{z} = \sum_{t,i}  c^{b}_{z,t,i} + \sum_{t,i} c^{e}_{z,t,i} + \sum_{i} c^{om}_{z,i} + c^{g}_{z} + 
 c^{r}_{z} + c^{s,v}_{z} + c^{x}_{z} + c^{q}_{z} \qquad \qquad \forall z
%\end{split}
\end{align}
The components of zonal system costs are calculated as given in equations (\ref{fuel_cost}) to (\ref{lost_load_cost}).
Lower-case $c$ represent total cost, while upper-case $C$ denotes specific, annualized capital cost of technology investment. Prices for fuels and \ce{CO2} are denoted by $P$.
\begin{align}
&c^{b}_{z,t,i}& =&\ \sum_{f} \left( P_{t,z,f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i \label{fuel_cost} \\
&c^{e}_{z,t,i}& =&\ \sum_{f} \left( P^{e}_{t,z} \: \varepsilon_{f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i\\
&c^{om}_{z,i}& =&\ O^{q}_{i} \left(\bar{G}_{z,i} - \bar{g}^{-}_{z,i} + \bar{g}^{+}_{z,i}\right) + \sum_{t,m} \left(O^{v}_{i} \: g_{z,t,i,m}\right) \qquad \qquad &\forall z,i \\
&c^{g}_{z}& =&\ C^{g}_{z,i} \: \bar{g}^{+}_{z,i} \qquad \qquad &\forall z\\
&c^{r}_{z}& =&\ C^{r}_{z,n} \: \bar{r}^{+}_{z,n} \qquad \qquad &\forall z\\
&c^{s,v}_{z}& =&\ C^{s}_{z,k} \: \bar{s}^{+}_{z,k} + C^{v}_{z,k} \: v^{+}_{z,k} \qquad \qquad &\forall z\\
&c^{x}_{z}& =&\ \frac{1}{2} \: \sum_{zz} (C^{x}_{z} \: \delta_{z,zz} \: \bar{x}^{+}_{z,zz}) \qquad \qquad &\forall z \label{transmission_expansion_cost}\\
&c^{q}_{z}& =&\ \mu \sum_{t,m} q^{-}_{z,t,m} \qquad \qquad &\forall z \label{lost_load_cost}
\end{align}

>>>>

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\DeclareMathOperator*{\argmax}{arg\,max}

@@

\begin{longtable}{p{3.6cm} c c p{6cm}}
%\caption{Sets} \\ 
\hline
\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{elements} \\
\hline \hline
fuels & $f \in F$ & \texttt{f} & \texttt{nuclear, lignite, coal, gas, oil, biomass, power} \\ \hline
power generation technologies & $i \in I$ & \texttt{i} & 
\texttt{nuc, lig\_stm, lig\_stm\_chp, lig\_boa, lig\_boa\_chp,coal\_sub, coal\_sub\_chp, coal\_sc, coal\_sc\_chp, coal\_usc, coal\_usc\_chp, coal\_igcc, ng\_stm, 
ng\_stm\_chp, ng\_ctb\_lo, ng\_ctb\_lo\_chp, ng\_ctb\_hi, ng\_ctb\_hi\_chp, ng\_cc\_lo, ng\_cc\_lo\_chp, ng\_cc\_hi, ng\_cc\_hi\_chp, ng\_mtr, ng\_mtr\_chp, 
ng\_boiler\_chp, oil\_stm, oil\_stm\_chp, oil\_ctb, oil\_ctb\_chp, oil\_cc, oil\_cc\_chp, bio, bio\_chp, heatpump\_pth} \\ \hline
power to heat technologies & $h \in H \subset I$ & \texttt{h(i)} & \texttt{heatpump\_pth}  \\ \hline
CHP technologies & $j \in J \subset I$ & \texttt{j(i)} & \texttt{lig\_stm\_chp, lig\_boa\_chp, coal\_sub\_chp, coal\_sc\_chp, coal\_usc\_chp, ng\_stm\_chp, 
ng\_ctb\_lo\_chp, ng\_ctb\_hi\_chp, ng\_cc\_lo\_chp, ng\_cc\_hi\_chp, ng\_mtr\_chp, ng\_boiler\_chp, oil\_stm\_chp, oil\_ctb\_chp, oil\_cc\_chp, bio\_chp} \\ \hline
storage technologies & $k \in K$ & \texttt{k} & \texttt{res\_day, res\_week, res\_season, psp\_day, psp\_week, psp\_season, battery} \\ \hline
feasible operation region limits & $l \in L$ & \texttt{l} & \texttt{l1, l2, l3, l4} \\ \hline
energy products & $m \in M$ & \texttt{m} & \texttt{el, ht} \\ \hline
intermittent generators & $n \in N$ & \texttt{n} & \texttt{wind\_on, wind\_off, pv, ror}\\ \hline
time periods (hours) & $t \in T$ & \texttt{t} & \texttt{t1, t2, \ldots, t8760} \\ \hline
market zones & $z \in Z$ & \texttt{z} & \texttt{AT, DE} \\ \hline \hline
\end{longtable}

@@

\begin{longtable}{p{4.1cm} c l c}
\hline
	\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{unit} \\ \hline \hline
% LOWER-CASE GREEK LETTERS
	distance between countries & $\delta_{z,zz}$ & \texttt{DISTANCE(z,zz)} & km \\ \hline
	fuel emission intensity & $\varepsilon_{f}$ & \texttt{CO2\_INTENSITY(f)} & $\text{t}_{\ce{CO2}}$ / MWh \\ \hline
	power plant efficiency & $\eta_{i,m,f}$ & \texttt{EFFICIENCY\_G(i,m,f)} & MWh / MWh \\ \hline
	efficiency power out & $\eta^{out}_{z,k}$ & \makecell[l]{\texttt{EFFICIENCY\_S\_OUT(k)}} &  \\ \hline
	efficiency power in & $\eta^{in}_{z,k}$ & \makecell[l]{\texttt{EFFICIENCY\_S\_IN(k)}} &  \\ \hline	
	scaling factor for peak load & $\lambda_{z}$ & \texttt{LAMBDA(z)} &  \\ \hline
	value of lost load & $\mu_{z}$ & \texttt{VALUE\_NSE(z)} & \EUR / MWh \\ \hline
	reservoir inflows & $\rho_{z,t,k}$ & \makecell[l]{\texttt{INFLOWS(z,t,k)}} & MW \\ \hline	
	scaling factor for peak intermittent generation & $\sigma_{z}$ & \texttt{SIGMA(z)} &  \\ \hline
	intermittent generation profile & $\phi_{z,t,n}$ & \texttt{GEN\_PROFILE(z,t,n)} & $[0,1]$ \\ \hline
	peak intermittent generation profile & $\widehat{\phi}_{z,n}$ & \texttt{PEAK\_PROFILE(z,n)} & $[0,1]$ \\ \hline	
	inputs of feasible operating region & $\chi_{i,l,f}$ & \texttt{FEASIBLE\_INPUT(i,l,f)} & $[0,1]$ \\ \hline
	output tuples of feasible operating region & $\psi_{i,l,m}$ & \texttt{FEASIBLE\_OUTPUT(i,l,m)} & $[0,1]$ \\ \hline
% UPPER-CASE LATIN LETTERS	
	capital cost of intermittent generators (specific, annuity) & $C^{r}_{z,n}$ & \texttt{CAPITALCOST\_R(z,n)} & k\EUR / GW \\ \hline
	capital cost of thermal generators (specific, annuity) & $C^{g}_{z,i}$ & \texttt{CAPITALCOST\_G(z,i)} & k\EUR / GW \\ \hline
	capital cost of storages - power (specific, annuity) & $C^{s}_{z,k}$ & \makecell[l]{\texttt{CAPITALCOST\_S(z,k)}} & k\EUR / GW \\ \hline
	capital cost of storages - energy (specific, annuity) & $C^{v}_{z,k}$ & \makecell[l]{\texttt{CAPITALCOST\_V(z,k)}} & k\EUR / GW \\ \hline
	capital cost of transmission capacity & $C^{x}$ & \texttt{CAPITALCOST\_X} & k\EUR / GW \\ \hline
	energy demand & $D_{z,t,m}$ & \texttt{DEMAND(z,t,m)} & GW \\ \hline
	peak demand & $\widehat{D}_{z,m}$ & \texttt{PEAK\_LOAD(z,m)} & GW \\ \hline
	initial capacity of dispatchable generators & $\widetilde{G}_{z,i}$ & \texttt{INITIAL\_CAP\_G(z,tec)} & GW \\ \hline
\pagebreak
%\caption[Parameters]{(continued)}\\ 
\hline
	% UPPER-CASE LATIN LETTERS (CONTINUED)
	\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{unit} \\ \hline \hline
	quasi-fixed O\&M cost & $O^{q}_{i}$ & \texttt{OM\_COST\_QFIX(i)} & k\EUR / GW \\ \hline
	variable O\&M cost & $O^{v}_{i}$ & \texttt{OM\_COST\_VAR(i)} & \EUR / MWh \\ \hline
	\ce{CO2} price & $P^{e}_{t,z}$ & \texttt{PRICE\_CO2(t,z)} & \EUR / $\text{t}_{\ce{CO2}}$ \\ \hline
	fuel price & $P_{t,z,f}$ & \texttt{PRICE\_FUEL(t,z,f)} & \EUR / MWh \\ \hline
	initial capacity of intermittent generators & $\widetilde{R}_{z,n}$ & \texttt{INITIAL\_CAP\_R(z,n)} & GW \\ \hline
	max power out & $\widetilde{S}^{out}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_S\_OUT(z,k)}} & GW \\ \hline
	max power in & $\widetilde{S}^{in}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_S\_IN(z,k)}} & GW \\ \hline
	max energy stored & $\widetilde{V}_{z,k}$ & \makecell[l]{\texttt{INITIAL\_CAP\_V(z,k)}} &  \\ \hline
	installed available transmission capacity & $\widetilde{X}_{z,zz}$ & \texttt{INITIAL\_CAP\_X(z,zz)} & GW \\ \hline \hline	
\end{longtable}

@@

\begin{longtable}{p{4.1cm} c l c}
%\caption{Variables}\label{variables} \\ 
\hline
\textbf{name} & \makecell[l]{\textbf{math} \\ \textbf{symbol}} & \makecell[l]{\textbf{GAMS} \\\textbf{symbol}} & \textbf{Unit} \\ \hline \hline
fuel burn for energy generation & $b_{z,t,i,f}$ & \texttt{b(z,t,i,f)} & GW \\ \hline
total system cost & $c$ & \texttt{cost\_system} & k\EUR \\ \hline
zonal system cost & $c_{z}$ & \texttt{cost\_zonal(z)} & k\EUR \\ \hline
fuel cost & $c^{b}_{z,t,i}$ & \texttt{cost\_fuel(z,t,i)} & k\EUR \\ \hline
emission cost & $c^{e}_{z,t,i}$ & \texttt{cost\_co2(z,t,i)} & k\EUR \\ \hline
total o\&m cost & $c^{om}_{z,i}$ & \texttt{cost\_om(z,i)} & k\EUR \\ \hline
capital cost of generators & $c^{g}_{z}$ & \texttt{cost\_invest\_g(z)} & k\EUR \\ \hline
total cost of non-served load & $c^{q}_{z}$ & \texttt{cost\_nse(z)} & k\EUR \\ \hline
capital cost of intermittent generators & $c^{r}_{z}$ & \texttt{cost\_invest\_r(z)} & k\EUR \\ \hline
capital cost of storages & $c^{s,v}_{z}$ & \texttt{cost\_invest\_sv(z)} & k\EUR \\ \hline
capital cost of interconnectors & $c^{x}_{z}$ & \texttt{cost\_invest\_x(z)} & k\EUR \\ \hline
\ce{CO2} emissions & $e_{z}$ & \texttt{emission\_co2(z)} & t \ce{CO2} \\ \hline
added capacity of dispatchables & $\widetilde{g}^{+}_{z,i}$ & \texttt{add\_g(z,i)} & GW \\ \hline
decommissioned capacity of dispatchables & $\widetilde{g}^{-}_{z,i}$ & \texttt{deco\_g(z,i)} & GW \\ \hline
energy generated by conventionals & $g_{z,t,i,m}$ & \texttt{g(z,t,i,m)} & GW \\ \hline
curtailed energy & $q^{+}_{z,t}$ & \texttt{q\_curtail(z,t)} & GW \\ \hline
non-served energy & $q^{-}_{z,t,m}$ & \texttt{q\_nse(z,t,m)} & GW \\ \hline
added capacity of intermittents & $\widetilde{r}^{+}_{z,n}$ & \texttt{add\_r(z,n)} & GW \\ \hline
electricity generated by intermittents & $r_{z,t,n}$ & \texttt{r(z,t,n)} & GW \\ \hline
added storage capacity (power) & $\widetilde{s}^{+}_{z,k}$ & \texttt{add\_s(z,k)} & GW \\ \hline
energy stored in & $s^{in}_{z,t,k}$ & \texttt{s\_in(z,t,k)} & GW \\ \hline
energy stored out & $s^{out}_{z,t,k}$ & \texttt{s\_out(z,t,k)} & GW \\ \hline
added storage capacity (energy) & $\widetilde{v}^{+}_{z,k}$ & \texttt{add\_v(z,k)} & GWh \\ \hline
storage energy content & $v_{z,t,k}$ & \texttt{v(z,t,k)} & GWh \\ \hline
operating region weight & $w_{z,t,i,l}$ & \texttt{w(z,t,i,l)} & \\ \hline
added transmission capacity & $\widetilde{x}^{+}_{z,zz}$ & \texttt{add\_x(z,zz)} & GW \\ \hline
electricity net export & $x_{z,zz,t}$ & \texttt{x(z,zz,t)} & GW \\ \hline \hline
\end{longtable}

@@

\section{Naming system}
\ctable[
	maxwidth = 160mm,
	pos = h]
	{X p{2.0cm} p{2.0cm} p{2.0cm} p{2.0cm} p{2.0cm}}
	{
	\tnote[$\dagger$]{parameter} 
	\tnote[$\ddagger$]{variable}
	}
	{\FL 
	 &  initial capacity\tmark[$\dagger$] & added capacity\tmark[$\ddagger$] & \makecell[l]{decom-\\missioned \\capacity\tmark[$\ddagger$]} & \makecell[l]{specific \\investment \\cost\tmark[$\dagger$]} & dispatch\tmark[$\ddagger$] \ML
	thermal units & $\widetilde{G}_{z,i}$ & $\widetilde{g}^{+}_{z,i}$ & $\widetilde{g}^{-}_{z,i}$ & $C^{g}_{z,i}$ & $g_{z,t,i,m}$ \ML
	\makecell[l]{intermittent \\units} &  $\widetilde{R}_{z,n}$ & $\widetilde{r}^{+}_{z,n}$ & $\widetilde{r}^{-}_{z,n}$ & $C^{r}_{z,n}$ & $r_{z,t,n}$ \ML
	\makecell[l]{storages \\(power)} & $\widetilde{S}_{z,k}$ & $\widetilde{s}^{+}_{z,k}$ & $\widetilde{s}^{-}_{z,k}$ & $C^{s}_{z,k}$ & $s_{z,t,k}$ \ML
	\makecell[l]{storages \\(energy)} & $\widetilde{V}_{z,k}$ & $\widetilde{v}^{+}_{z,k}$ & $\widetilde{v}^{-}_{z,t,k}$ & $C^{v}_{z,k}$ & na \ML
	transmission &  $\widetilde{X}_{z,zz}$ & $\widetilde{x}^{+}_{z,zz}$ & $\widetilde{x}^{-}_{z,zz}$ & $C^{x}_{z,zz}$ & $x_{z,zz,t}$ \LL
}

@@

\paragraph{Model objective}
\emph{medea} minimizes total system cost $c$, i.e. the total cost of generating electricity and heat from technologies and capacities adequate to meet demand, over a large number  of decision variables, essentially representing investment and dispatch decisions in each market zone $z$ of the modelled energy systems.
\begin{align}
\min c = \sum_{z} (c_{z})
\end{align}
Zonal system costs $c_{z}$ are the sum of fuel cost $c^{b}_{z,t,i}$, emission cost $c^{e}_{z,t,i}$, operation and maintenance cost $c^{om}_{z,i}$, capital costs of investment in conventional and intermittent generation ($c^{g}_{z}$, $c^{r}_{z}$), storage ($c^{s,v}_{z}$) and transmission ($c^{x}_{z}$) equipment, and the cost of non-served load ($c^{q}_{z}$) that accrues when demand is not met, e.g. when there is a power outage. 
\begin{align}
%\begin{split}
c_{z} = \sum_{t,i}  c^{b}_{z,t,i} + \sum_{t,i} c^{e}_{z,t,i} + \sum_{i} c^{om}_{z,i} + c^{g}_{z} + 
 c^{r}_{z} + c^{s,v}_{z} + c^{x}_{z} + c^{q}_{z} \qquad \qquad \forall z
%\end{split}
\end{align}
The components of zonal system costs are calculated as given in equations (\ref{fuel_cost}) to (\ref{lost_load_cost}).
Lower-case $c$ represent total cost, while upper-case $C$ denotes specific, annualized capital cost of technology investment. Prices for fuels and \ce{CO2} are denoted by $P$.
\begin{align}
&c^{b}_{z,t,i}& =&\ \sum_{f} \left( P_{t,z,f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i \label{fuel_cost} \\
&c^{e}_{z,t,i}& =&\ \sum_{f} \left( P^{e}_{t,z} \: \varepsilon_{f} \: b_{t,z,i,f} \right) \qquad \qquad &\forall z,t,i\\
&c^{om}_{z,i}& =&\ O^{q}_{i} \left(\widetilde{G}_{z,i} - \widetilde{g}^{-}_{z,i} + \widetilde{g}^{+}_{z,i}\right) + \sum_{t,m} \left(O^{v}_{i} \: g_{z,t,i,m}\right) \qquad \qquad &\forall z,i \\
&c^{g}_{z}& =&\ \sum_{i} \left( C^{g}_{z,i} \: \widetilde{g}^{+}_{z,i} \right) \qquad \qquad &\forall z\\
&c^{r}_{z}& =&\ \sum_{n} \left( C^{r}_{z,n} \: \widetilde{r}^{+}_{z,n} \right) \qquad \qquad &\forall z\\
&c^{s,v}_{z}& =&\ \sum_{k} \left( C^{s}_{z,k} \: \widetilde{s}^{+}_{z,k} + C^{v}_{z,k} \: v^{+}_{z,k} \right) \qquad \qquad &\forall z\\
&c^{x}_{z}& =&\ \frac{1}{2} \: \sum_{zz} (C^{x} \: \delta_{z,zz} \: \widetilde{x}^{+}_{z,zz}) \qquad \qquad &\forall z \label{transmission_expansion_cost}\\
&c^{q}_{z}& =&\ \mu \sum_{t,m} q^{-}_{z,t,m} \qquad \qquad &\forall z \label{lost_load_cost}
\end{align}
